---
source: Rmd
title: "3 - I - Remixing datasets based on uniform variable names"
output: html_document
teaching: 30
exercises: 2
questions:
- "Which properties help us merge datasets?"
objectives:
- "Finding both data and software for re-use."
- "Visualising data from two different sets."
keypoints:
- "The FAIR-er a dataset, the easier its re-use in answering a new research questions"
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
opts_chunk$set(fig.path = "../fig/01-")
```

We want to analyse ice core data from two different regions of the world (and projects):

- [Arctic (NGRIP) `10.1594/PANGAEA.586886`](https://doi.pangaea.de/10.1594/PANGAEA.586886)
- [Antarctic (DomeC) `10.1594/PANGAEA.57629`](https://doi.pangaea.de/10.1594/PANGAEA.57629)

Please skim over both pages.

---

- Noticed the DOIs? Both datasets are **F**indable :-)
- Even more so, because both are annotated with geographic and event info.
- How are they licensed? Under what conditions are we allowed to use the data?
- Noticed the `Download` links over `https`? **A**ccessible data :-)


Now, will the data also be **i**nteroperable and **r**eusable on the technical level? The list of parameters in both cases show that `Age` and `d18O H2O` were measured. Sounds good, but will it blend???

> ## Challenge: What should be our first move towards our goal of analysing the datasets?
>
> Of course, we need to download the datasets somehow. We could:
>
> - `Download dataset as tab-delimited text` manually, saving the files
> - write our own little download function that could for example take a vector dataset IDs (`c(586886, 57629)`) and somehow
>
> Which one do you prefer?
>
> > ## Solution
> >
> > Neither ;-) There is a third option: looking for an R package
> > related to the data repository. Search
> > [CRAN.R-project.org/web/packages/available_packages_by_name.html
> > ](https://cran.r-project.org/web/packages/available_packages_by_name.html)
> >  or [https://ropensci.org/packages/](ROpenSci.org/packages) for
> > "PANGAEA".
> {: .solution}
{: .challenge}


> ## Challenge: Before installing your search result, check whether it seems useful.
>
> How would you go about this in case of an R-package or Python module?
>
> > ## Solution:
> >
> > [ROpenSci.github.io/pangaear/reference
> > ](https://ropensci.github.io/pangaear/reference/index.html) shows
> > an overview of all functions. `pg_data()` sounds like what we need.
> {: .solution}
{: .challenge}


```{r download, warning=FALSE, message=FALSE}
# install.packages("pangaear")
library(pangaear)
NGRIP <- pg_data(doi = "10.1594/PANGAEA.57629")
DomeC <- pg_data("10.1594/PANGAEA.586886")
```

Before analysing any data, we should get an overview of the R objects we created just now by the downloads:

```{r str1}
str(NGRIP)
```

```{r str2}
str(DomeC)
```

Both lists contain some metadata and the actual `data` as a `tbl_df`. What is that? [Hint: It's from the tidyverse](https://tibble.tidyverse.org/).

To verify that both variables `Age` and `d18O H2O` are really labelled in exactly the same way, we extract the variable names of both tibbles and throw out others that exist only in one tibble.


```{r vars}
DomeC <- DomeC[[1]]$data
NGRIP <- NGRIP[[1]]$data
intersect(names(DomeC), names(NGRIP))
```

This is great! We can now plot both datasets' `d18O H2O` values against the same `Age` axis. Because the variables names contain spaces and brackets, we use the "back tick" character (\`) around both.

```{r plot}
library(ggplot2)
ggplot(data = DomeC,
       mapping = aes(x = `Age [ka BP]`, y = `d18O H2O [per mil SMOW]`)) +
  geom_point() + # inherits data & mapping from above
  geom_point(data = NGRIP, # overwrites above data, but inherits x & y
             color = "dark green") +
  scale_x_reverse() # because Age means the past
```

Incidentally, the `Dome C` core (black, east Antarctica) captured higher `d18O` concentrations, than `NGRIP` (*Green*land).

Integrating the two datasets with this little code was possible, only because both variables were named exactly the same and thus easily machine-readable. If they hadn't been, we would have needed to:

a) either rename one between downloading and plotting, or

```{r str_diff, eval=FALSE}
names(NGRIP$`some other Age variable's name`) <- names(DomeC$`Age [ka BP]`)
names(NGRIP$`some other d18O variable's name`) <- names(DomeC$`d18O H2O [per mil SMOW]`)
```

b) specify in `geom_point` which second `y` should be plotted.

```{r plot2, eval=FALSE}
ggplot(data = DomeC,
       mapping = aes(x = `Age [ka BP]`, y = `d18O H2O [per mil SMOW]`)) +
  geom_point() +
  geom_point(data = NGRIP,
             mapping = aes(x = `some other Age variable's name`,
                           y = `some other d18O variable's name`)
             color = "dark green")
```

So, this seemed pretty FAIR, right? Granted, **i**nteroperability encompasses a [few qualities](https://blogs.tib.eu/wp/tib/2017/09/12/the-fair-data-principles-for-research-data/#i) besides uniform variable names. However, the datasets were well **f**indable & **a**ccessible, community-standard variable naming and the CC-BY-3.0 license meant mean good **r**eusability.

This is the power of FAIR combined with nicely packaged-up R code: Reducing the burden of finding, downloading and cleaning up datasets.
